---
title: "Izzy's Final Project"
author: "Izzy Grabowski"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## IDEAS (DELETE THIS)

Where I got the data from: https://catalog.data.gov/dataset/electric-vehicle-population-data
https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?n=PET&s=EMM_EPM0_PTE_SWA_DPG&f=M 

-make several ggplots
-clean data
- CREATE NEW variables using the 'mutate' function

1. When and where are EVs most popular?
Which counties and cities in Washington have the highest number of registered EVs?
Are rural or urban areas more likely to adopt EVs?

2. What types of electric vehicles are being adopted?
What is the distribution of BEVs vs. PHEVs by county or city?

3. How do economic or policy factors relate to EV adoption?

4. What is the electric range distribution, and how does it vary?
What are the most common electric ranges?

5. Are there geographic clusters of EV use related to utility providers?
Do certain Electric Utilities have more EV customers?
Are these utilities investing more in infrastructure, like charging stations?
Is EV adoption correlated with specific utility regions?

6. How does vehicle type correlate with census tracts?
Are EVs more common in specific 2020 Census Tracts?



## Introduction



In this project, I have explored data about electric vehicles. 



```{r}
##CLEANING

library(readr)
Electric_Vehicle_Population_Data <- read_csv("~/Desktop/Electric_Vehicle_Population_Data.csv")
#View(Electric_Vehicle_Population_Data)

colnames(Electric_Vehicle_Population_Data)

summary(Electric_Vehicle_Population_Data)

ev_data <- Electric_Vehicle_Population_Data

str(ev_data)
summary(ev_data)
head(ev_data)


ev_data <- na.omit(ev_data)
colSums(is.na(ev_data)) # remove rows with lots of missing data


ev_data$City <- tolower(ev_data$City)
ev_data$Make <- toupper(ev_data$Make) # this makes text in the data consistent


ev_data <- ev_data[!duplicated(ev_data), ] # getting rid of duplicates


```


## Top 10 Counties and Cities by EV Registrations
-Which counties and cities in Washington have the highest number of registered EVs?



```{r}

library(dplyr)
library(ggplot2)

# Group by County and count
top_counties <- ev_data %>%
  group_by(County) %>%
  summarise(EV_Count = n()) %>%
  arrange(desc(EV_Count)) %>%
  slice_max(EV_Count, n = 10)  # Top 10 counties

# Plot
ggplot(top_counties, aes(x = reorder(County, EV_Count), y = EV_Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Counties in Washington by EV Registrations",
       x = "County", y = "Number of EVs")

# Group by City and count
top_cities <- ev_data %>%
  group_by(City) %>%
  summarise(EV_Count = n()) %>%
  arrange(desc(EV_Count)) %>%
  slice_max(EV_Count, n = 10)  # Top 10 cities

# Plot
ggplot(top_cities, aes(x = reorder(City, EV_Count), y = EV_Count)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Top 10 Cities in Washington by EV Registrations",
       x = "City", y = "Number of EVs")


```

Based on these plots, Seattle and King county have the largest number of electric vehicles. Seattle is in King county so these plots match the data. 

# When and where are EVs most popular?


```{r}
library(dplyr)
library(ggplot2)

# Group by Model Year
ev_trend <- ev_data %>%
  group_by(`Model Year`) %>%
  summarise(EV_Count = n())

# Plot
ggplot(ev_trend, aes(x = `Model Year`, y = EV_Count)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "EV Registrations by Model Year",
       x = "Model Year", y = "Number of EVs") +
  theme_minimal()


# Group by Model Year and County
ev_year_location <- ev_data %>%
  group_by(`Model Year`, County) %>%
  summarise(EV_Count = n())

# Plot: use facet to compare counties over time
ggplot(ev_year_location, aes(x = `Model Year`, y = EV_Count)) +
  geom_line() +
  facet_wrap(~ County, scales = "free_y") +
  labs(title = "EV Popularity Over Time by County",
       x = "Model Year", y = "Number of EVs") +
  theme_minimal()


ev_peak <- ev_data %>%
  group_by(County, `Model Year`) %>%
  summarise(EV_Count = n()) %>%
  arrange(desc(EV_Count)) %>%
  slice(1)

print(ev_peak)

```

It looks like that the peak EV registrations was in 2023, which steeply trended down in 2025. This could possibly be because prices went up significantly, or many people already had electric vehicles and the demand lowered. 



Are rural or urban areas more likely to adopt EVs?



```{r}
unique(ev_data$County) # figure out which counties are rural and which ones are urban from this site: https://doh.wa.gov/sites/default/files/legacy/Documents/Pubs/609003.pdf



urban_counties <- c("Whatcom", "Snohomish", "Kitsap", "King", "Pierce", "Thurston", "Clark", "Benton", "Spokane")

# Add a new column: Urban vs Rural
ev_data <- ev_data %>%
  mutate(Area_Type = ifelse(County %in% urban_counties, "Urban", "Rural"))

# Count EVs by area type
area_summary <- ev_data %>%
  group_by(Area_Type) %>%
  summarise(EV_Count = n())

# Plot
ggplot(area_summary, aes(x = Area_Type, y = EV_Count, fill = Area_Type)) +
  geom_bar(stat = "identity") +
  labs(title = "EV Adoption: Urban vs Rural Areas",
       x = "Area Type", y = "Number of EVs") +
  theme_minimal()




```
From this result, it is clear that electric vehicles are more wide spread in urban counties. This makes sense because urban areas can be more environmentally friendly. Urban areas, especially cities, value efficiency in resources is some cases. So, it makes sense more people within cities would have electric vehicles, there also is probably more access to charging stations in urban areas. 


## 2

What types of electric vehicles are being adopted?

```{r}

library(dplyr)
library(ggplot2)

# Count of each EV type
ev_type_summary <- ev_data %>%
  group_by(`Electric Vehicle Type`) %>%
  summarise(EV_Count = n())

# Bar plot
ggplot(ev_type_summary, aes(x = reorder(`Electric Vehicle Type`, -EV_Count), y = EV_Count, fill = `Electric Vehicle Type`)) +
  geom_bar(stat = "identity") +
  labs(title = "Types of Electric Vehicles Adopted in Washington",
       x = "Electric Vehicle Type", y = "Number of EVs") +
  theme_minimal()


```
Clearly, battery electric vehicles are more common in WA compared to plug-in hybrid electric vehicles. This is better for the environment since hybrid cars combine a gasoline engine with an electric battery. So, more battery electric vehicles reduces the carbon footprint more. 

What is the distribution of BEVs vs. PHEVs by county or city?

```{r}

# Count EV types by County
ev_by_type_and_county <- ev_data %>%
  group_by(County, `Electric Vehicle Type`) %>%
  summarise(EV_Count = n())

# Stacked bar chart
ggplot(ev_by_type_and_county, aes(x = reorder(County, -EV_Count), y = EV_Count, fill = `Electric Vehicle Type`)) +
  geom_bar(stat = "identity") +
  labs(title = "BEV vs. PHEV Distribution by County",
       x = "County", y = "Number of EVs") +
  theme_minimal() +
  coord_flip()



```
This distribution is reflecting how the urban counties have a higher number of battery only electric vehicles compared to hybrid vehicles. For each county, the ratios are very similar when comparing the counts. 



## 3


How do economic or policy factors relate to EV adoption?
Here, I am using data on retail gas prices in Washington every month. 


```{r}
library(readr)
library(dplyr)
library(tidyr)

# Read the CSV file and skip the first four rows of metadata
Gas_Prices <- read_csv("~/Documents/Washington_All_Grades_All_Formulations_Retail_Gasoline_Prices.csv", skip = 4)

head(Gas_Prices)

#str(Gas_Prices)
#View(Gas_Prices)
#colnames(Gas_Prices)



```

```{r}


library(lubridate)
library(ggplot2)

gas_yearly <- Gas_Prices %>%
  rename(Month = `Month`, Price = `Washington All Grades All Formulations Retail Gasoline Prices Dollars per Gallon`) %>%
  mutate(
    Date = parse_date_time(Month, orders = "b Y"),
    Year = year(Date),
    Price = as.numeric(Price)
  ) %>%
  group_by(Year) %>%
  summarise(Avg_Gas_Price = mean(Price, na.rm = TRUE))

ev_yearly <- ev_data %>%
  group_by(`Model Year`) %>%
  summarise(EV_Count = n()) %>%
  rename(Year = `Model Year`)

ev_gas_combined <- left_join(ev_yearly, gas_yearly, by = "Year")

ggplot(ev_gas_combined, aes(x = Avg_Gas_Price, y = EV_Count)) +
  geom_point(color = "darkred", size = 3) +
  labs(title = "EV Count by Model Year vs. Average Gas Price",
       x = "Average Gas Price (USD)",
       y = "EV Registrations (by Model Year)") +
  theme_minimal()


cor.test(ev_gas_combined$Avg_Gas_Price, ev_gas_combined$EV_Count)
#testing correlation between the two

```

This result is very interesting because when gas prices were below $3, EV registration counts did not increase nor decrease. Once gas prices began to increase and stay between 3-4 dollars, Ev registrations slightly increased. Finally, when gas prices exceeded 4 dollars, EV registrations in Washington increased significantly. 


```{r}
ev_type_year <- ev_data %>%
  group_by(`Model Year`, `Electric Vehicle Type`) %>%
  summarise(EV_Count = n()) %>%
  rename(Year = `Model Year`)


ev_type_gas <- left_join(ev_type_year, gas_yearly, by = "Year")


library(ggplot2)

ggplot(ev_type_gas, aes(x = Avg_Gas_Price, y = EV_Count, color = `Electric Vehicle Type`)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "EV Type Registrations vs. Gas Prices",
       x = "Average Gas Price (USD)",
       y = "EV Registrations",
       color = "EV Type") +
  theme_minimal()



```


```{r}
correlations <- ev_type_gas %>%
  group_by(`Electric Vehicle Type`) %>%
  summarise(Correlation = cor(Avg_Gas_Price, EV_Count, use = "complete.obs"))

print(correlations)

```
