---
title: "Izzy's Final Project"
author: "Izzy Grabowski"
date: "2025-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## IDEAS (DELETE THIS)

Where I got the data from: https://catalog.data.gov/dataset/electric-vehicle-population-data

-make several ggplots
-clean data
- CREATE NEW variables using the 'mutate' function

1. When and where are EVs most popular?
Which counties and cities in Washington have the highest number of registered EVs?
Are rural or urban areas more likely to adopt EVs?

2. What types of electric vehicles are being adopted?
What is the distribution of BEVs vs. PHEVs by county or city?

3. How do economic or policy factors relate to EV adoption?

4. What is the electric range distribution, and how does it vary?
What are the most common electric ranges?

5. Are there geographic clusters of EV use related to utility providers?
Do certain Electric Utilities have more EV customers?
Are these utilities investing more in infrastructure, like charging stations?
Is EV adoption correlated with specific utility regions?

6. How does vehicle type correlate with census tracts?
Are EVs more common in specific 2020 Census Tracts?



## Introduction



In this project, I have explored data about electric vehicles. 



```{r}
##CLEANING

library(readr)
Electric_Vehicle_Population_Data <- read_csv("~/Desktop/Electric_Vehicle_Population_Data.csv")
#View(Electric_Vehicle_Population_Data)

colnames(Electric_Vehicle_Population_Data)

summary(Electric_Vehicle_Population_Data)

ev_data <- Electric_Vehicle_Population_Data

str(ev_data)
summary(ev_data)
head(ev_data)


ev_data <- na.omit(ev_data)
colSums(is.na(ev_data)) # remove rows with lots of missing data


ev_data$City <- tolower(ev_data$City)
ev_data$Make <- toupper(ev_data$Make) # this makes text in the data consistent


ev_data <- ev_data[!duplicated(ev_data), ] # getting rid of duplicates


```

